---
title: "Isoforma  Vignette"
author: "Logan Lewis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# What is Isoforma?

Isoforma is a package for quantifying positional isomers in MS2 spectra data. 
Currently, analysis of this type of data requires the use of several, separate 
tools, which is inconvenient and time-consuming for researchers. This goal of this
software is offer all the functionality needed for this analysis and provide it in
a streamlined package. The data is given to the user as a table of relative 
percentages as well as an abundance matrix.

Much of the backend functionality is drawn from the pspecterlib package, including
generating metadata objects. More information about the backend package
can be found [here](https://github.com/EMSL-Computing/pspecterlib).

# Wrapper Function - isoforma_pipeline()

For user convenience, we offer a wrapper function, 'Isoforma_pipeline()' which
contains the necessary functions to complete the Isoforma workflow. The function
requires the unmodified peptide sequence, modification (see create_modifications_object 
for details of formatting), retention time start and end, and a path to a .mzML file.

#TWO SECTIONS

Below is an example modified ubiquitin data.

```{r}
#What do I do about this? I don't have data in this format

tmpdir <- tempdir()

isoforma_pipeline(unmodified_sequence = "LQIFVKTLTGKTITLEVEPSDTIENVKAKIQDKEGIPPDQQRLIFAGKQLEDGRTLSDYNIQKESTLHLVLRLRGG",
                  modifications = "Aviett number**,V(17,26,70)[1]",
                  rt_start, 
                  rt_end,
                  mzML_path,
                  output_path = tmpdir,
                  mass_window = 10,
                  mz_round = 3,
                  correlation_score = 0.7)
```

# Accessory Functions

## create_modifications_object()

### WELL REPLACE WITH PSPECTER FUNCTION

Isoforma requires peptide strings in a specific format. This function will output
a string in the correct format needed for the package. The user must provide a 
modification name that is in the PTM database, the possible positions of the modification
and the number of modification.
**Clarify symbols**

```{r}
# Single modification of a single PTM examples 
create_modifications_object("TRICITIES", "Methyl,I(3,5,7)[1]")


# Multiple modifications of a single PTM example
create_modifications_object("TRICITIES", "Methyl,I(3,5,7)[2]") 

# Multiple modifications with a fixed position
create_modifications_object("TRICITIES", "Methyl,I(3^,5,7)[2]") 

# Multiple modifications with two fixed positions 
create_modifications_object("TRICITIES", "Methyl,I(3^,5,7^)[2]")

# Multiple modifications of a single PTM with any "X" residue
create_modifications_object("TRICITIES", "Methyl,X(1,2,3,4,5,6,7,8,9)[2]")

# Multiple modifications with multiple PTMs examples 
create_modifications_object("TRICITIES", "Methyl,X(1)[1];Acetyl,X(2,4,9)[1]")
create_modifications_object("TRICITIES", "Methyl,X(1,2,3,4,5,6,7,8,9)[1];Acetyl,X(2,4,9)[1]")
create_modifications_object("TRICITIES", "Methyl,X(1^,2,3,4,5,6,7^,8,9)[3];Acetyl,X(2,4,9)[1]")
```

## write_mgf_simple()

This function will write a .tsv output of a given spectra. This can be useful to
users who wish to see their data in this format has they are running the pipeline.

```{r}
write_mgf_simple(round(SumSpectra$`M/Z`, 4), round(SumSpectra$Intensity, 4), file.path(out, "SumSpectra.mgf"),
                   title = strsplit(out, "/", fixed = T) %>% unlist() %>% tail(1),
                   pepmass = ExactMass, charge = paste0(max(ScanNumbers$`Precursor Charge`), "+"), 
                   rt = mean(ScanNumbers$`Retention Time`), scans = paste0(ScanNumbers$`Scan Number`, collapse = " & "))
```

## ptm_heatmap()

This function creates a heatmap based on the amount of each post-transcriptional 
modification in the data. This offers a useful visualizatoin of the data to users.

```{r}
ptm_heatmap(FragmentsPerPTM)
```

## annotated_spectrum_ptms_plot

This returns a [trelliscope](https://github.com/hafen/trelliscopejs) display of spectra data. 
Trelliscope offers a great interactive UI that gives a plethora of options for exploring and
visualizing their data.

```{r}
annotated_spectrum_ptms_plot(SumSpectra, FragmentsPerPTM)
```

# Pipeline Functions

## pull_scan_numbers()

#WILL BE SKIPPED IF YOU PROVIDE A SPECTRA

This function generates scan numbers based on a Precursor Mass, mass tolerance, and 
isolation window filtering. This can be as a vector or as a dataframe if AsDataframe
is set to TRUE.

```{r}
#*ASK DAVID FOR THIS DATA!

# Generate the ScanMetadata object 
ScanMetadata <- pspecterlib::get_scan_metadata(
   MSPath = "/Users/degn400/Desktop/IsoForma_Test/Sorghum-Histone0622162L11.mzML"
)
 
 # Run function
pull_scan_numbers(Sequence = "M.SGRGKGGKGLGKGGAKRHRKVLRDNIQGITKPAIRRLARRGGVKRISGLIYEETRGVLKIFLENVIRDAVTYTEHARRKTVTAMDVVYALKRQGRTLYGFGG",
                   Modification = "Acetyl,X(1*,5,8,12,16,20)[2]",
                   ScanMetdata = ScanMetadata,
                   PrecursorMass = 11355.38,
                   RTStart = 30,
                   RTEnd = 60)
```

## sum_spectra()

#WILL BE SKIPPED IF YOU PROVIDE A SPECTRA
#REPLACE SUMMING FUNCTION WITH ISOMATCH? IF RESULTS DONT CHANGE

Users can sum together spectra based on scan number using this function. Summing
spectra help gives stronger signal by combining the results of several replicates.

```{r}
#*ASK DAVID FOR THIS DATA!

# Generate the ScanMetadata object 
ScanMetadata <- pspecterlib::get_scan_metadata(
    MSPath = "/Users/degn400/Desktop/IsoForma_Test/Sorghum-Histone0622162L11.mzML"
)

# Run function
sum_spectra(ScanMetdata = ScanMetadata,
            ScanNumbers = c(1236, 1237, 1240, 1316, 1497, 1520, 1534))
```



## fragments_per_ptm()

This will calculate the number of fragments of each post-transcriptional modification
in the data, and matches the peaks using the matching alogrithm from the pspecter 
package.The function returns a matched_peaks_isoforma object.

```{r}
#FIX WITH DATA
FragmentsPerPTM <- fragments_per_ptm(Sequences = c(unmodified_sequence, create_modifications_object(unmodified_sequence, modifications)),
                                       SummedSpectra = SumSpectra,
                                       PrecursorCharge = PrecursorCharge,
                                       ActivationMethod = ActivationMethod,
                                       CorrelationScore = 0.7,
                                       IsotopeAlgorithm = IsotopeAlgorithm,
                                       PPMThreshold = 5,
                                       Messages = TRUE)
```

## sum_isotopes()

This step in the pipeline sums all isotopes and charge states of a fragment (i.e. 
a “c” ion that is an M+1 isotope with charge 5 will be summed to an M isotope with 
charge 6, etc). Returns a summed_isotopes_isoforma object.

```{r}
SummedIsotopes <- sum_isotopes(FragmentsPerPTM)
```

## abundance_matrix()

This function creates an abundance matrix where rows are ion type for every effected 
residue, and columns are the summed PTM amounts based on the inputed summed_isotopes_isoforma 
object. The output is assigned the class "abundance_matrix_isoforma", but takes 
the form of a dataframe and can be saved to a .csv using utils::write.csv(). If one ion doen't produce expected results, try another one based on activation method.

```{r}
AbundanceMatrix <- abundance_matrix("c", SummedIsotopes)
```

## calculate_proportions()

Here the proportions of each of the fragments are calculated using the previously
generated abundance matrix. It is calculated as as the abundance for a PTM over 
the max abundance in that row. This function returns a dataframe as well as an optional
graphical output, providing a visual representation of the fragment proportions.

```{r}
Proportions <- calculate_proportions(AbundanceMatrix, modifications, IncludePlot = TRUE)
```

